sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-03-14 doc/paper.Rmd
fix typo in owl caption
JL comments

---
title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanographic and Atmospheric Administration, Seattle, WA, USA\n|
    Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{lscape}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyze these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but can be extended to penalized likelihood or Bayesian inference. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
    | Keywords:
      Gaussian Markov random fields (GMRF),
      generalized linear mixed effects models (GLMM),
      INLA,
      SPDE,
      species distribution modelling,
      R package,
      spatio-temporal or spatial-temporal,
      Template Model Builder
---

@@

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart, due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper, and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit GRFs to large spatial datasets [@lindgren2011].

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid latitudes of North America in much greater numbers in some winters than others.
The reasons for this interannual variation in the number of individuals migrating south are not well understood, but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.

@@

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NOA unit.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA will find inlabru approachable.
Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed-wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.

@@

Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-gaussian random fields [e.g., @anderson2019].
The included TMB `.cpp` file also provides a high-quality, bug-free, and tested model template that can be modified to add additional features.
<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend.
<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
<!-- Line 275: I would suggest adding a MAJOR benefit of sdmTMB for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates! This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it. It saves me a bunch of time. If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it. -->

@@

sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages.
sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages.
Penalized spline support was possible thanks to mgcv [@wood2017a].
We thank the authors of all these packages.
We thank M. Lindmark, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

>>>>

---
title: "sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields"
author: "|
  Sean C. Anderson$^1$^[Corresponding author: sean.anderson@dfo-mpo.gc.ca],
  Eric J. Ward$^2$,
  Philina A. English$^1$,
  Lewis A. K. Barnett$^3$\n|
  $^1$Pacific Biological Station, Fisheries and Oceans Canada, Nanaimo, BC, Canada\n|
  $^2$Northwest Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
  $^3$Alaska Fisheries Science Center, National Marine Fisheries Service,\n|
      National Oceanic and Atmospheric Administration, Seattle, WA, USA\n|
    Running headline: sdmTMB: spatial and spatiotemporal GLMMs\n"
output:
    # bookdown::word_document2:
    bookdown::pdf_document2:
      toc: false
      number_sections: false
      fig_caption: true
      highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
link-citations: yes
header-includes:
  \usepackage{setspace}\onehalfspacing
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \usepackage{tablefootnote}
  \linenumbers
  \newcommand{\beginsupplement}{
  \setcounter{equation}{0}
  \renewcommand{\theequation}{S.\arabic{equation}}
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
  \usepackage{lscape}
  \widowpenalty10000
  \clubpenalty10000
editor_options:
  markdown:
    wrap: sentence
abstract: |
    | \noindent 1. Geostatistical data---spatially referenced observations related to some continuous spatial phenomenon---are ubiquitous in ecology and can reveal ecological processes and inform management decisions. However, appropriate models to analyze these data, such as generalized linear mixed effects models (GLMMs) with Gaussian random fields, are often computationally intensive and challenging to implement, interpret, and evaluate.
    | 2. Here, we introduce the R package sdmTMB, which implements predictive-process SPDE- (stochastic partial differential equation) based spatial and spatiotemporal models. Estimation is conducted via maximum marginal likelihood with Template Model Builder (TMB) but can be extended to penalized likelihood or Bayesian inference. We describe the statistical model, illustrate the package's use through two case studies, and compare the functionality, speed, and interface to related software.
    | 3. We highlight advantages of using sdmTMB for this class of models: (1) sdmTMB provides a flexible interface familiar to users of `glm()`, lme4, glmmTMB, or mgcv; (2) estimation is often faster than alternatives; (3) sdmTMB provides simple out-of-sample cross validation; (4) non-stationary processes (time-varying and  spatially varying coefficients) are easily constructed with a formula interface; and (5) sdmTMB includes features not available as a combination in related packages (e.g., penalized smoothers and break-point effects, anisotropy, abundance index standardization).
    | 4. We hope that sdmTMB's user-friendly interface will open this useful class of models to a wider audience within species distribution modelling and beyond.
    | Keywords:
      Gaussian Markov random fields (GMRF),
      generalized linear mixed effects models (GLMM),
      INLA,
      SPDE,
      species distribution modelling,
      R package,
      spatio-temporal or spatial-temporal,
      Template Model Builder
---

@@

Ecological data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about ecological processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of ecological interest itself by, for example, identifying locations with similar population responses [e.g., @thorson2019d; @barnett2021].
<!-- LB: seems like we needed some explicit example of why it would be of ecological interest itself to those who are unfamiliar, so I added allusion to local trends...but maybe also mentioning quantification of patchiness or something similar?. More generally, it might be good to expand on the utility of such models for applied ecology, if we can find room within the tight word limit? -->

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the INLA R package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in TMB [Template Model Builder, @kristensen2016] that partially relies on INLA to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

@@

Snowy Owls (*Bubo scandiacus*) breed on the arctic tundra and are irruptive migrants, meaning that they appear across the mid-latitudes of North America in much greater numbers in some winters than others.
The reasons for this interannual variation in the number of individuals migrating south are not well understood but seem to be related to high abundances of food during the breeding season and therefore sharp increases in breeding ground population densities [@robillard2016].
The North Atlantic Oscillation Index (NAO) has been linked to productivity of both owls and their prey in Europe [@millon2014].
Because both productivity and the choice of wintering location could be influenced by climate, we tested for a spatially varying effect of annual mean NAO index on winter abundance across the southern boundary of their winter distribution.
We fit counts observed in North America during annual Christmas Bird Counts [@cbc] using a negative binomial (NB2) distribution, random intercepts for year, spatial and spatiotemporal random fields, and a spatially varying coefficient associated with the NAO.

@@

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US.
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8).
The effect is multiplicative on owl count per NAO unit.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface and speed.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA will find inlabru approachable.
Users familiar with mgcv can adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications (Appendix 5), users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed-wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022].
INLA/inlabru, or spaMM for spatial models, scale the best with very high-resolution meshes.

@@

Additional functionality in sdmTMB not already mentioned includes interpolating across missing time slices and forecasting, the barrier SPDE model of @bakka2019, time-varying spatiotemporal covariance parameters, and simulation from the parameter joint precision matrix.
Future development may include zero-inflated models, improvements to Bayesian sampling efficiency [e.g., @monnahan2021], continuous time models [e.g., @blangiardo2015book], and non-Gaussian random fields [e.g., @anderson2019].
The included TMB `.cpp` file also provides a high-quality and tested model template that can be modified to add additional features.
<!-- Not sure what kind of refs you want here?  INLA's non-gaussian random fields are mentioned here @bakka2018review-->
We may introduce explicit delta models, but many such models can already be fit with sdmTMB by modelling the two components separately.
We may introduce capability for multivariate response data, but VAST already focuses on such models with a TMB backend.
<!-- Furthermore, we would only add this functionality if we could maintain a straightforward interface for univariate data. -->
<!-- Line 275: I would suggest adding a MAJOR benefit of sdmTMB for advanced users: a tool for quickly building high-quality, bug-free, and tested model templates! This is by far my favourite feature of the brms package, as it allows me to build a pretty complex ‘base-model’ in c++/stan code before adding my own custom functionality on top of it. It saves me a bunch of time. If you don’t currently have a ‘return_TMB_code=TRUE’ option, then I would highly recommend it. -->

@@

sdmTMB would not be possible without the TMB [@kristensen2016] and INLA [@rue2009; @lindgren2011; @lindgren2015] R packages.
sdmTMB is heavily inspired by and in some places code has been adapted from both the VAST [@thorson2019] and glmmTMB [@brooks2017] R packages.
Penalized spline support was possible thanks to mgcv [@wood2017a].
We thank the authors of all these packages.
We thank M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
